Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-04-06 paper/tidy-missing-data-paper.Rmd
notes from meeting with Di

- re-arrange this paragraph around

To explore missing data there needs to be a robust representation of missing
values that integrates into analysis pipelines. This means a data structure that is easily transported, and not a special class reliant on attributes that can be lost in analysis. The $B$ matrix, where 0 and 1 indicate not missing and missing (respectively), provides a useful template, as it is very common in missing data literature.  This matrix was used to explore missing values in
the interactive graphics library XGobi, called a "missing value
shadow", or "shadow matrix", defined as a copy of the original data with
indicator values of missingness. The shadow matrix could be interactively linked to the data.

There are some limitations to the shadow matrix, however. Namely, the values 0 and 1 can be confusing representations of missing values - does
0 indicate an absence of observation, or the presence of a missing value?

@@

We propose a new form for tidy representation of missing data based on these ideas from past research. Four features are added to the shadow matrix to facilitate analysis of missing data, illustrated in Figure \@ref(fig:shadow-matrix-progress)

@@

1. **Missing value labels**. Simple labels, "NA" and "!NA", for missing and not missing, clearly identifies missing values for analysis and plotting. This is an improvement over the commonly used binary matrix ($B$), which makes it unclear if 0 indicates the absence (0) of observation, or the presence (1) of a missing value. Missing value labels can be represented as "factor" types values, with text values "NA" and "!NA" and underlying number values 1 and 0, for missing and not missing (Figure \@ref(fig:shadow-matrix-progress) parts 1-3).

(principle: clarity of labelling - transparent what the matrix means, because these could equally be "missing" or "present" - for this. Anybody looking at these values)

1. **Special missing values**: Building on **missing value labels**, the values in the shadow matrix can be "special" missing values, indicated by "NA_\<suffix\>" - for example: `NA_instrument` uses a short label, "instrument", indicating instrument error resulting in missing values. These could be also used to indicate imputations, (Table \@ref(tab:shadow-encoding)). The underlying number representation changes from binary to integer, allowing for unlimited missingness features. For example, instrument failure and drop out would be represented as 2 and 3 (Table \@ref(tab:shadow-encoding)). Encoding special missing values is achieved by defining logical conditions and suffixes with the `recode_shadow` function in \pkg{naniar} (\@ref(verbs-recode)).

Other statistical programs (SPSS, SAS, and Stata) represent missingness values as a full-stop, `.`, and allow for recording special missing values as `.a` - `.z`.

SAS, SPSS, and STATA all implement

The special values from these languages break the rule of "keeping one thing in a cell", as they record both the value and the multivariate missingness state. Additionally, special missing values in other statistical programming languages are encoded at a lower level in the fundamental `NA` type, which is not possible in R.

(discuss the principles, rather than what we've done)

(in the same column you have mixed information, which breaks the tidy principle of one column having one type of value)

@@

Visualizing missingness in a dataset is essential to understand its structure. This section discusses four different areas of visualization of missing data: overviews (\@ref(overview-graphics)), univariate (\@ref(univariate-graphics)), bivariate (\@ref(bivariate-graphics)), and multivariate (\@ref(multivariate-graphics)). All plots are created using \pkg{ggplot2}, giving users clearer control over the plot appearance.

## Overview graphics {#overview-graphics}

@@

## Univariate graphics {#univariate-graphics}

@@

## Bivariate graphics {#bivariate-graphics}

@@

## Multivariate graphics {#multivariate-graphics}

>>>>

To explore missing data there needs to be a robust representation of missing
values that integrates into analysis pipelines. This means a data structure that is easily transported, and not a special class reliant on attributes that can be lost in analysis. The $B$ matrix, where 0 and 1 indicate not missing and missing (respectively), provides a useful template, as it is very common in missing data literature.  This matrix was used to explore missing values in
the interactive graphics library XGobi, called a "missing value
shadow", or "shadow matrix", defined as a copy of the original data with
indicator values of missingness. The shadow matrix could be interactively linked to the data. There are some limitations to the shadow matrix, however. Namely, the values 0 and 1 can be confusing representations of missing values - does 0 indicate an absence of observation, or the presence of a missing value?

@@

We propose a new form for tidy representation of missing data based on these ideas from past research. Four features are added to the shadow matrix to facilitate analysis of missing data, illustrated in Figure \@ref(fig:shadow-matrix-progress).

@@

1. **Missing value labels**. Simple labels, "NA" and "!NA", for missing and not missing, clearly identifies missing values for analysis and plotting. This is an improvement over the commonly used binary matrix ($B$), which makes it unclear if 0 indicates the absence (0) of observation, or the presence (1) of a missing value. Missing value labels can be represented as "factor" types values, with text values "NA" and "!NA" and underlying number values 1 and 0, for missing and not missing (Figure \@ref(fig:shadow-matrix-progress) parts 1-3). (principle: clarity of labelling - transparent what the matrix means, because these could equally be "missing" or "present" - for this. Anybody looking at these values)

1. **Special missing values**: Building on **missing value labels**, the values in the shadow matrix can be "special" missing values, indicated by "NA_\<suffix\>" - for example: `NA_instrument` uses a short label, "instrument", indicating instrument error resulting in missing values. These could be also used to indicate imputations, (Table \@ref(tab:shadow-encoding)). The underlying number representation changes from binary to integer, allowing for unlimited missingness features. For example, instrument failure and drop out would be represented as 2 and 3 (Table \@ref(tab:shadow-encoding)). Encoding special missing values is achieved by defining logical conditions and suffixes with the `recode_shadow` function in \pkg{naniar} (\@ref(verbs-recode)). {Other statistical programs (SPSS, SAS, and Stata) represent missingness values as a full-stop, `.`, and allow for recording special missing values as `.a` - `.z`.  SAS, SPSS, and STATA all implement } {The special values from these languages break the rule of "keeping one thing in a cell", as they record both the value and the multivariate missingness state. Additionally, special missing values in other statistical programming languages are encoded at a lower level in the fundamental `NA` type, which is not possible in R.} {(discuss the principles, rather than what we've done)}. {(in the same column you have mixed information, which breaks the tidy principle of one column having one type of value)}.

@@

note: First sentence needs to be more on track with the overall summary of what this section is about...

- Graphics are important to understand the missingness structure.
- The multivariate relationships
- Working through a hierarchy of how to explore missings - from overview to 
- Sentence linking the two.
- How to visualise a binary matrix of missing vs not, and the dependencies between those binary variables as well.
- It is more than that, when we visualise the complete cases, we don't want to lose track of the missings themselves. This comes back to the data structure
- gg_miss_var and gg_miss_upset and vis_miss are all showing how we can summarise the missings.
- but the addition of missing counts into a histogram, allows us to understand that there are many missings - and we have a skewed distribution o Ozone. But we want to know where these values might have been located. tying this back to the nabular structure, and also the terminology of MCAR MAR, etc. (We would expect if data were MCAR for these points to be representatively thoughout the histogram)

When we define the graphics, describe which part of the nabular matrix we are going to display (overview - just binary parts), the summaries, are summaries of the binary, the histogram and points are both the missings and the complete values.
-

- Combining that with the data itself, looking at the patterns in the complete cases, 
- from the overview through to the univariate

Visualizing missingness in a dataset is essential to understand its structure. This section discusses four different areas of visualization of missing data: overviews (\@ref(overviews)), univariate (\@ref(univariate)), bivariate (\@ref(bivariate)), and multivariate (\@ref(multivariate)). All plots are created using \pkg{ggplot2}, giving users clearer control over the plot appearance.

## Overviews {#overviews}

@@

## Univariate {#univariate}

@@

## Bivariate {#bivariate}

@@

## Multivariate {#multivariate}
